`forestecology` package for modeling interspecies competition between trees

### 2021-01-09 etc/paper/forestecology.Rmd
cleaned manuscript

@@

# Load spatial packages:
# devtools::install_github("rvalavi/blockCV")
library(blockCV)
library(sf)
library(sfheaders)

@@

```{r growth-histogram, out.width="100%", fig.cap="Distribution of average annual growth in DBH for both sites."}
# Both Big Woods & SCBI
bind_rows(
  bw_growth_df %>% st_drop_geometry() %>% select(growth) %>% mutate(Site = "Big Woods"),
  scbi_growth_df %>% st_drop_geometry() %>% select(growth) %>% mutate(Site = "SCBI")
) %>% 
  ggplot(aes(x = growth, y = ..density.., fill = Site)) +
  geom_histogram(alpha = 0.5, position = "identity", binwidth = 0.05) +
  labs(x = "Average annual growth in dbh (cm per yr)") +
  coord_cartesian(xlim = c(-0.5, 1))
```

@@

```{r bw-define-cv-folds}
set.seed(76)
bw_spatialBlock <- spatialBlock(
  speciesData = bw_growth_df, theRange = cv_fold_size,
  k = 28, xOffset = 0.5, yOffset = 0,
  verbose = FALSE
)

# Add foldID to each tree
bw_growth_df <- bw_growth_df %>%
  mutate(foldID = bw_spatialBlock$foldID)

# Visualize grid. Why does fold 19 repeat?
bw_spatialBlock$plots +
  geom_sf(data = bw_growth_df %>% sample_frac(0.2), aes(col = factor(foldID)), size = 0.1)

# Remove empty folds
bw_growth_df <- bw_growth_df %>%
  filter(!foldID %in% c(19, 23, 21, 17, 8, 19)) %>%
  mutate(foldID = as.character(foldID))
```

@@

```{r}
fold1 <- rbind(c(0, 300), c(150, 300), c(150, 600), c(0, 600), c(0, 300)) %>%
  sf_polygon() %>% 
  mutate(folds = 1)
fold2 <- rbind(c(150, 300), c(300, 300), c(300, 600), c(150, 600), c(150, 300)) %>%
  sf_polygon() %>% 
  mutate(folds = 2)
scbi_cv_grid <- bind_rows(fold1, fold2)
```

@@

```{r}
head(focal_vs_comp_bw)
```

@@

```{r}
head(focal_vs_comp_scbi)
```

@@

Next we fit the following linear model to the dbh of each focal tree. Let $i = 1, \ldots, n_j$ index all $n_j$ trees of ``focal'' species group $j$; let $j = 1, \ldots, J$ index all $J$ focal species groups; and let $k = 1, \ldots, K$ index all $K$ ``competitor'' species groups. We modeled the growth in diameter per year $y_{ij}$ (in centimeters per year) of the $i^{th}$ tree of focal species group $j$ as a linear model $f$ of the following covariates $\vec{x}_{ij}$

@@

```{r scbi-model-residuals, out.width="100%", fig.cap="Spatial distribution of residuals for model applied to SCBI data."}
ggplot(scbi_growth_df_noCV, aes(x = growth, y = growth_hat)) +
  geom_point(size = 0.5, color = rgb(0, 0, 0, 0.25)) +
  stat_smooth(method = "lm") +
  geom_abline(slope = 1, intercept = 0) +
  coord_fixed() +
  labs(
    x = "Observed growth in dbh", y = "Predicted growth in dbh",
    title = "Predicted vs Observed Growth"
  )
```

```{r scbi-model-residuals-2, out.width="100%", fig.cap="Spatial distribution of residuals for model applied to SCBI data part 2."}
scbi_growth_df_noCV %>%
  st_as_sf() %>%
  # TODO: Need to investigate missingness
  filter(!is.na(growth_hat)) %>%
  mutate(
    error = growth - growth_hat,
    error_bin = cut_number(error, n = 5),
    error_compress = ifelse(error < -0.75, -0.75, ifelse(error > 0.75, 0.75, error))
  ) %>%
  ggplot() +
  geom_sf(aes(col = error_compress), size = 1) +
  theme_bw() +
  scale_color_gradient2(
    low = "#ef8a62", mid = "#f7f7f7", high = "#67a9cf",
    name = expression(paste("Residual (cm ", y^{-1}, ")")),
    breaks = seq(from = -0.75, to = 0.75, by = 0.25),
    labels = c("< -0.75", "-0.5", "0.25", "0", "0.25", "0.5", "> 0.75")
  ) +
  labs(x = "Meter", y = "Meter")
```

@@

Now the posterior parameter outputs of `fit_bayesian_model()` are passed to `plot_posterior_parameters()` to generate visualizations of the posterior parameters. These visualizations are displayed in Figure 5 of @allen_permutation_2020. For simplicity we only plot a subset of the species families.

```{r bw-posterior-viz-1}
posterior_plots <- plot_posterior_parameters(
  posterior_param = posterior_param_bw,
  sp_to_plot = c("cornaceae", "fagaceae", "hamamelidaceae", "juglandaceae", 
                 "lauraceae", "rosaceae", "sapindaceae", "ulmaceae")
)
```

@@

```{r bw-posterior-viz-beta0, out.width="50%", fig.cap="Posterior distribution of beta0."}
posterior_plots[["beta_0"]]
```

@@

```{r bw-posterior-viz-beta-dbh, out.width="100%", fig.cap="Posterior distribution of betadbh."}
posterior_plots[["beta_dbh"]]
```

@@

```{r bw-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda."}
posterior_plots[["lambda"]]
```

@@

```{r scbi-posterior-viz-1}
posterior_plots_scbi <- plot_posterior_parameters(
  posterior_param = posterior_param_scbi,
  sp_to_plot = c("quru", "litu", "cagl", "cato")
)
```

```{r scbi-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda."}
posterior_plots_scbi[["lambda"]]
```

>>>>

The code that generates Figures are included in the supplementary materials.

@@

# Load spatial packages:
library(blockCV)
library(sf)
library(sfheaders)

@@

```{r growth-histogram, out.width="100%", fig.cap="Distribution of average annual growth in DBH for both sites.", echo = FALSE}
# Both Big Woods & SCBI
bind_rows(
  bw_growth_df %>% st_drop_geometry() %>% select(growth) %>% mutate(Site = "Big Woods"),
  scbi_growth_df %>% st_drop_geometry() %>% select(growth) %>% mutate(Site = "SCBI")
) %>% 
  ggplot(aes(x = growth, y = ..density.., fill = Site)) +
  geom_histogram(alpha = 0.5, position = "identity", binwidth = 0.05) +
  labs(x = "Average annual growth in dbh (cm per yr)") +
  coord_cartesian(xlim = c(-0.5, 1))
```

@@

```{r bw-define-cv-folds, out.width="100%", fig.cap="blockCV function output."}
set.seed(76)
bw_spatialBlock <- spatialBlock(
  speciesData = bw_growth_df, theRange = cv_fold_size, k = 28, xOffset = 0.5, 
  yOffset = 0, verbose = FALSE, showBlocks = FALSE
)
```

# Add foldID to each tree
```{r}
bw_growth_df <- bw_growth_df %>%
  mutate(foldID = bw_spatialBlock$foldID)
```

```{r bw-define-cv-folds-with-trees, out.width="100%", fig.cap="Inspect blocks closely."}
# Visualize grid. Why does fold 19 repeat?
ggplot() +
  geom_sf(data = bw_spatialBlock$blocks %>% st_as_sf()) + 
  geom_sf(data = bw_growth_df %>% sample_frac(0.2), 
          aes(col = factor(foldID)), size = 0.1) + 
  theme(legend.position="bottom")
```

# Remove empty folds

```{r}
bw_growth_df <- bw_growth_df %>%
  filter(!foldID %in% c(19, 23, 21, 17, 8, 19)) %>%
  mutate(foldID = as.character(foldID))
```

@@

```{r}
fold1 <- rbind(c(0, 300), c(150, 300), c(150, 600), c(0, 600)) %>%
  sf_polygon() %>% 
  mutate(folds = 1)
fold2 <- rbind(c(150, 300), c(300, 300), c(300, 600), c(150, 600)) %>%
  sf_polygon() %>% 
  mutate(folds = 2)
scbi_cv_grid <- bind_rows(fold1, fold2)
```

@@

@@

@@

Next we fit the following linear model to the dbh of each focal tree. Let $i = 1, \ldots, n_j$ index all $n_j$ trees of "focal" species group $j$; let $j = 1, \ldots, J$ index all $J$ focal species groups; and let $k = 1, \ldots, K$ index all $K$ "competitor" species groups. We modeled the growth in diameter per year $y_{ij}$ (in centimeters per year) of the $i^{th}$ tree of focal species group $j$ as a linear model $f$ of the following covariates $\vec{x}_{ij}$

@@

```{r scbi-model-residuals, out.width="100%", fig.cap="Predicted versus observed growth.", echo = FALSE}
ggplot(scbi_growth_df_noCV, aes(x = growth, y = growth_hat)) +
  geom_point(size = 0.5, color = rgb(0, 0, 0, 0.25)) +
  stat_smooth(method = "lm") +
  geom_abline(slope = 1, intercept = 0) +
  coord_fixed() +
  labs(
    x = "Observed growth in dbh", y = "Predicted growth in dbh",
    title = "Predicted vs Observed Growth"
  )
```

```{r scbi-model-residuals-2, out.width="80%", fig.cap="Spatial distribution of residuals for model applied to SCBI data.", echo = FALSE}
scbi_growth_df_noCV %>%
  st_as_sf() %>%
  # TODO: Need to investigate missingness
  filter(!is.na(growth_hat)) %>%
  mutate(
    error = growth - growth_hat,
    error_bin = cut_number(error, n = 5),
    error_compress = ifelse(error < -0.75, -0.75, ifelse(error > 0.75, 0.75, error))
  ) %>%
  ggplot() +
  geom_sf(aes(col = error_compress), size = 1) +
  theme_bw() +
  scale_color_gradient2(
    low = "#ef8a62", mid = "#f7f7f7", high = "#67a9cf",
    name = expression(paste("Residual (cm ", y^{-1}, ")")),
    breaks = seq(from = -0.75, to = 0.75, by = 0.25),
    labels = c("< -0.75", "-0.5", "0.25", "0", "0.25", "0.5", "> 0.75")
  ) +
  labs(x = "Meter", y = "Meter")
```

@@

Now the posterior parameter outputs of `fit_bayesian_model()` are passed to `plot_bayesian_model_parameters()` to generate visualizations of the posterior parameters. These visualizations are displayed in Figure 5 of @allen_permutation_2020. For simplicity we only plot a subset of the species families.

```{r bw-posterior-viz-1}
posterior_plots_bw <- plot_bayesian_model_parameters(
  posterior_param = posterior_param_bw,
  sp_to_plot = c("cornaceae", "fagaceae", "hamamelidaceae", "juglandaceae", 
                 "lauraceae", "rosaceae", "sapindaceae", "ulmaceae")
)
```

@@

```{r bw-posterior-viz-beta0, out.width="100%", fig.cap="Posterior distribution of beta0."}
posterior_plots_bw[["beta_0"]]
```

@@

```{r bw-posterior-viz-beta-dbh, out.width="100%", fig.cap="Posterior distribution of betadbh."}
posterior_plots_bw[["beta_dbh"]]
```

@@

```{r bw-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda's for Big Woods."}
posterior_plots_bw[["lambda"]]
```

@@

```{r scbi-posterior-viz-1}
posterior_plots_scbi <- plot_bayesian_model_parameters(
  posterior_param = posterior_param_scbi,
  sp_to_plot = c("quru", "litu", "cagl", "cato")
)
```

```{r scbi-posterior-viz-lambda, out.width="100%", fig.cap="Posterior distribution of lambda's for SCBI."}
posterior_plots_scbi[["lambda"]]
```
